---
title: "HUDM6122 Homework_02"
author: "Chenguang Pan"
date: "2023-02-06"
output:
  pdf_document:
    toc: false
    toc_depth: 4
    number_sections: true
    keep_tex: true
    highlight: tango
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(
  cache = TRUE,
  prompt = TRUE,
  comment = '',
  collapse = TRUE,
  warning = FALSE,
  message = FALSE)
```

## Ex. 2.1  
*Use the bivariate boxplot on the scatterplot of each pair of variables in the air pollution data to identify any outliers. Calculate the correlation between each pair of variables using all the data and the data with any identified outliers removed. Comment on the results.*

**MY SOLUTION:**  

Several techniques worth to be noted:
- use a for-loop within a for-loop to map all pairs
- use `text()` to give each point a name.
- the bivariate boxplot function `bvbox` is inclueded in the package `MVA`

```{r, fig.show='hold',out.width="33%",out.height= "25%"}
# import the data
library(HSAUR2)
library(MVA)
attach(USairpollution)
head(USairpollution)
dim(USairpollution)
# draw the bivariate boxplot of each pair of variables with a for-loop
for (i in 1:7) {
  for (j in i:7) {
    if (i != j) {
      var_pair <- USairpollution[, c(i, j)]
      bvbox(var_pair, 
            xlab = names(USairpollution)[i],
            ylab = names(USairpollution)[j])
      text(USairpollution[,i],USairpollution[,j],
          row.names(USairpollution), # to add the point name
          pos=2,col = "blue")
    } 
  }
}
```    
From the graphs, we can easily find that the outliers among the observations are "Chicago", "Detroit", "Cleveland","Philadelphia", "Miami", "Phoenix","Albuquerque", "Providence". I run the correlation matrix on all observations first.  
```{r}
# create the correaltion matrix on all observations
round(cor(USairpollution),2)
# remove all identified outliers
drop_city <- match(c("Chicago", "Detroit","Cleveland",
                     "Philadelphia", "Miami","Phoenix",
                     "Albuquerque", "Providence"), rownames(USairpollution))

round(cor(USairpollution[-drop_city,]),2)
```   
After dropping all the identified outliers, some of the correlation coefficients has changed to the opposite direction, like from positive to negative, others shrink or increase. It is reasonable since some outliers are with high leverage.

## Ex. 2.2  
*Compare the chi-plots with the corresponding scatterplots for each pair of variables in the air pollution data. Do you think that there is any advantage in the former?*  

**MY SOLUTION:**   
Several details should be noted.  
- For drawing many graphs in `Rmd` file with `knit`, it always reports error or `no such file or directory`. One can clear all the cache in R and cache file and Tex file in the file folder, and do not use the layout function `par()`.  
- The `chiplot` function is included in the package `MVA`. If two variables are independent, these value are asymptotically normal with mean zero; the xi values should show a non-systematic random fluctuation around zero.

```{r,fig.show='hold',out.width="33%",out.height= "25%"}
for (i in 1:7) {
  for (j in i:7) {
    if (i != j) {
      plot(USairpollution[,i],USairpollution[,j],
           xlab = names(USairpollution)[i], ylab=names(USairpollution)[j])
      chiplot(USairpollution[,i],USairpollution[,j],
              main = paste(names(USairpollution)[i],
                           "vs", 
                           names(USairpollution)[j]))
    } 
  }
}
```   
From the results, one can easily find that the scatter plots are sometimes difficult to identify the independence between two variables. But, comparatively the `chiplot` presents more straightforward way to tell this attribute. For example, it is hard to find the relation from the scattorplot for `manu` and `predays`, but the chiplot clearly demonstrates that these two varriables are independent.  

## Ex. 2.3  
*Construct a scatterplot matrix of the body measurements data that has the appropriate boxplot on the diagonal panels and bivariate boxplots on the other panels. Compare the plot with Figure 2.17, and say which diagram you find more informative about the data.*    

**MY SOLUTION:**   
The body measure dataset is not included in any packages. We need to create the dataset by ourselves. First, based on the codes offered by Prof. Motta, Giovanni, I created measure data
```{r}
# create the  body measure data. Codes offered by Prof.Motta.
measure <-
  structure(list(V1 = 1:20, V2 = c(34L, 37L, 38L, 36L, 38L, 43L,
                 40L, 38L, 40L, 41L, 36L, 36L, 34L, 33L, 36L, 37L, 34L, 36L, 38L,
                 35L), V3 = c(30L, 32L, 30L, 33L, 29L, 32L, 33L, 30L, 30L, 32L,
                 24L, 25L, 24L, 22L, 26L, 26L, 25L, 26L, 28L, 23L), V4 = c(32L,
                 37L, 36L, 39L, 33L, 38L, 42L, 40L, 37L, 39L, 35L, 37L, 37L, 34L,
                 38L, 37L, 38L, 37L, 40L, 35L)), .Names = c("V1", "V2", "V3",
                 "V4"), class = "data.frame", row.names = c(NA, -20L))
measure <- measure[,-1]
names(measure) <- c("chest", "waist", "hips")
measure$gender <- gl(2, 10)
levels(measure$gender) <- c("male", "female")

# to only extract the continuous data
measures <- measure[,c(1:3)]
par(mfrow=c(3, 3))
for (i in 1:3) {
  for (j in 1:ncol(measures)) {
    if(i != j) {
      bvbox(measures[, c(i, j)])
    }
    else {
      boxplot(measures[[i]])
    }
  }
}
```  
It is quite challenging to draw the graph in the author's way! First, `par("usr")` will return the coordiante of your current plot in the unit your data in `(xmin,xmax,ymin,ymax)` order.
```{r, include=FALSE}
# another way to draw more good-looking scatterplot matrix
panel.box <- function(x, ...)
{
  # use the par("usr") to get the coordinate of current plot
  usr <- par("usr");on.exit(par(usr));on.exit(par(usr))
  # reset the coordinate
  par(usr = c(usr[1:2], 0, 1.5) )
  boxplot(x)
}
# 
# pairs(measure,
#       diag.panel=panel.box,
#       panel = function(x,y) {
#         bvbox(cbind(x,y), add = T)})
#       
```  


## Ex. 2.4  
*Construct a further scatterplot matrix of the body measurements data that labels each point in a panel with the gender of the individual, and plot on each scatterplot the separate estimated bivariate densities for men and women.*    

**MY SOLUTION:**   

```{r,out.width="100%",out.height= "200%"}
ncols <- 3 # only the first 3 columns are numeric
par(mfrow=c(ncols, ncols))
for (i in 1:ncols) {
  for (j in 1:ncols) {
    plot(measure[, i], measure[, j], xlab = names(measure)[i], ylab=names(measure)[j])
    if(i != j) {
      bvbox(measure[measure$gender == "male", c(i, j)], add=TRUE)
      bvbox(measure[measure$gender == "female", c(i, j)], add=TRUE)
    }
    points(measure[measure$gender == "male", c(i, j)], col="blue")
    points(measure[measure$gender == "female", c(i, j)], col="red")
  }
}
```   

## Ex. 2.5  
*Construct a scatterplot matrix of the chemical composition of Romano-British pottery given in Chapter 1 (Table 1.3), identifying each unit by its kiln number and showing the estimated bivariate density on each panel. What does the resulting diagram tell you?*    

**MY SOLUTION:**   

```{r, eval=FALSE, include=FALSE}
ncols <- ncol(pottery)-1 # only the first 3 columns are numeric
par(mar = c(1, 1, 1, 1))
par(mfrow=c(ncols, ncols))
for (i in 1:ncols) {
  for (j in 1:ncols) {
    plot(pottery[, i], pottery[, j], xlab = names(pottery)[i], ylab=names(pottery)[j])
    if(i != j) {
      bvbox(pottery[pottery$kiln == 1, c(i, j)], add=TRUE)
      bvbox(pottery[pottery$kiln == 2, c(i, j)], add=TRUE)
      bvbox(pottery[pottery$kiln == 3, c(i, j)], add=TRUE)
      bvbox(pottery[pottery$kiln == 4, c(i, j)], add=TRUE)
      bvbox(pottery[pottery$kiln == 5, c(i, j)], add=TRUE)
    }
    points(pottery[pottery$kiln == 1, c(i, j)], col=1)
    points(pottery[pottery$kiln == 2, c(i, j)], col=2)
    points(pottery[pottery$kiln == 3, c(i, j)], col=3)
    points(pottery[pottery$kiln == 4, c(i, j)], col=4)
    points(pottery[pottery$kiln == 5, c(i, j)], col=5)
  }
}
```
```{r}
# pairs(measure, panel = function(x, y) plot(density(cbind(x, y)), add = TRUE))
library("KernSmooth")
par(mar = c(1, 1, 1, 1))
n = ncol(pottery) - 1
par(mfrow = c(n, n))
for (i in 1:n) {
  for(j in 1:n){
    
    if(i == j){
      #d = density(pottery[, i])
      #plot(d, main = colnames(pottery)[i])
      #text(d, cex = 0.6,
           #labels = abbreviate(pottery$kiln))
      par(mar = c(0,0,0,0))
      plot(c(0, 1), c(0, 1), ann = F, bty = 'n', type = 'n', xaxt = 'n', yaxt = 'n')
      
      text(x = 0.3, y = 0.5, paste(colnames(pottery)[i]), 
           cex = 1.5, col = "black", family="serif", font=2, adj=0.5)
    }
    else{
      measured = bkde2D(pottery[, c(i, j)], 
                bandwidth = sapply(pottery[, c(i, j)], dpik))
      
      plot(pottery[, c(i, j)],
           xlab = colnames(pottery)[i],
           ylab = colnames(pottery)[j])
      contour(x = measured$x1, y = measured$x2,
              z = measured$fhat, add = TRUE)
      text(pottery[,i], pottery[,j], cex = 0.6,
           labels = abbreviate(pottery$kiln))
      
    }
  }
  
}
```


## Ex. 2.6  
*Ex. 2.6 Construct a bubble plot of the earthquake data using latitude and longitude as the scatterplot and depth as the circles, with greater depths giving smaller circles. In addition, divide the magnitudes into three equal ranges and label the points in your bubble plot with a different symbol depending on the magnitude group into which the point falls.* 

```{r,eval=FALSE, include= FALSE,fig.show='hold',out.width="33%",out.height= "25%"}
library(lattice)
par(mfrow = c(1, 1))
#ylim <- with(quakes, range(long)) * c(0.95, 1)
sub1 = subset(quakes, subset = mag > 4 & mag <= 4.8)
sub2 = subset(quakes, subset = mag > 4.8 & mag <= 5.6)
sub3 = subset(quakes, subset = mag >5.6 & mag <= 6.4)

plot(sub1$lat, sub1$long, data = sub1,
     xlab = "Latitude",ylab = "Longitude", pch=10,
     )
with(sub1, symbols(lat, long, circles = 1000- depth,
                   inches = 0.3, add = T, col="red"))
plot(sub2$lat, sub2$long, data = sub2,
     xlab = "Latitude",ylab = "Longitude", pch=10,
     )
with(sub2, symbols(lat, long, circles = 1000- depth,
                   inches = 0.3, add = T, col="green"))
plot(sub3$lat, sub3$long, data = sub3,
     xlab = "Latitude",ylab = "Longitude", pch=10,
     )
with(sub3, symbols(lat, long, circles = 1000- depth,
                   inches = 0.3, add = T, col="blue"))
```

```{r}
library(lattice)
par(mfrow = c(1, 1))
#ylim <- with(quakes, range(long)) * c(0.95, 1)
sub1 = subset(quakes, subset = mag > 4 & mag <= 4.8)
sub2 = subset(quakes, subset = mag > 4.8 & mag <= 5.6)
sub3 = subset(quakes, subset = mag >5.6 & mag <= 6.4)


attach(quakes)
plot(lat, long,
     xlab = "Latitude",ylab = "Longitude", pch=10)

with(sub1, symbols(lat, long, circles = 1000- depth,
                   inches = 0.2, add = T, fg="green"))
with(sub2, symbols(lat, long, circles = 1000- depth,
                   inches = 0.2, add = T, fg="red"))
with(sub3, symbols(lat, long, circles = 1000- depth,
                   inches = 0.2, add = T, fg="blue"))
legend(-38,173, legend = c("4 < mag <= 4.8", 
                       "4.8 < mag <= 5.6",
                       "5.6 < mag <= 6.4"), fill=c("green","red","blue"))

```



