---
title: "HUDM6122 Homework_03"
author: "Chenguang Pan"
date: "2023-02-19"
output:
  pdf_document:
    toc: false
    toc_depth: 4
    number_sections: true
    keep_tex: true
    highlight: tango
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(
  cache = TRUE,
  prompt = TRUE,
  comment = '',
  collapse = TRUE,
  warning = FALSE,
  message = FALSE)
``` 

## Ex 3.1  
**MY SOLUTION:**  
```{r}
# import the package
library(MVA)
library(HSAUR2)
data(heptathlon)
# Create a scatterplot matrix with density contours
pairs(heptathlon[, -ncol(heptathlon)], upper.panel = function(x, y) {
  points(x, y)
  den <- MASS::kde2d(x, y)
  contour(den, add = TRUE, col = "red", lwd = 2)})
```  
Here, I use the `MASS::kde2d()` function to estimate the bivariate density of the data, and plot the contours of the density using the `contour()` function.

Comparing to the unenhanced scatter plot matrix, this mixed graph can help to easily find the specific characteristics of joint distribution of each pair, like the where is the center of the distribution.  

## Ex 3.2  
**MY SOLUTION:** 



