---
title: "HUDM6122 Homework_04"
author: "Chenguang Pan"
date: "2023-03-01"
output:
  pdf_document:
    toc: false
    toc_depth: 4
    number_sections: true
    keep_tex: true
    highlight: tango
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(
  cache = TRUE,
  prompt = TRUE,
  comment = '',
  collapse = TRUE,
  warning = FALSE,
  message = FALSE)
```   

## Ex 4.1  

*Consider 51 objects O1, . , O51 assumed to be arranged along a straight line with the jth object being located at a point with coordinate j. Define the similarity sij between object i and object j as...and then apply classical multidimensional scaling to the resulting dissimilarity matrix. Explain the shape of the derived two-dimensional solution.*  

**MY SOLUTION:**  
First, I define the function of dissimilarities $\delta_{ij}$ as follows, where the input paramter $X$ is a n*1 matrix that contains all the coordination of 51 objects.
```{r}
D_dis <- function(X) {
  n <- dim(X)[1] # to have the length of any input n*1 matrix
  S <- matrix(0, n, n) # to make a n*n empty matrix 
  D <- matrix(0, n, n) # to make a n*n empty matrix 
  # to make the similarity matrix by conditions
  for (i in c(1:n)) {
    for (j in c(1:n)){
      if (i == j){S[i,j] <- 9}
      else if (abs(i-j) >= 1 & abs(i-j) <= 3){S[i,j] <- 8}
      else if (abs(i-j) >= 4 & abs(i-j) <= 6){S[i,j] <- 7}
      else if (abs(i-j) >= 7 & abs(i-j) <= 9){S[i,j] <- 6}
      else if (abs(i-j) >= 10 & abs(i-j) <= 12){S[i,j] <- 5}
      else if (abs(i-j) >= 13 & abs(i-j) <= 15){S[i,j] <- 4}
      else if (abs(i-j) >= 16 & abs(i-j) <= 18){S[i,j] <- 3}
      else if (abs(i-j) >= 19 & abs(i-j) <= 21){S[i,j] <- 2}
      else if (abs(i-j) >= 22 & abs(i-j) <= 24){S[i,j] <- 1}
      else if (abs(i-j) >= 25){S[i,j] <- 0}
    }
  } # similarity matrix finished!
  # using the elements in the Similarity matrix to generate Dissimilarities Matrix
  for (i in c(1:n)) {
    for (j in c(1:n)) {
      D[i,j] <- sqrt(S[i,i] + S[j,j] - 2*S[i,j])
    }
  } # dissimilarity matrix finished!
  return(D)
}
```  
Next, I randomly generate a n*1 matrix with 51 integers by using `sample()` function. And plug this vector to the dissimilarity function above.  
```{r, eval=FALSE, include=FALSE}
# Note this code chunk won't run! and won't be displayed in the final PDF output!!
# just a record for wrong function for later reference!
obs_ <- function(n, # the number of objects you wanna generate
                     replace=TRUE, # randomly sample with replacement or not
                     start =1,  # range from start to end; 1-200 by default
                     end =200){
  number_vec <- sample(c(start:end),n,replace=replace)
  # change the number array to matrix
  return(matrix(number_vec,n,1))
}
```  
```{r}
obs_ <- function(n) {
  number_vec <- c(1:n)
  # change the number array to matrix
  return(matrix(number_vec,n,1))
}
``` 
The functions above looks good. I try to generate 51 observations and plug them into the dissimilarity matrix function to get the required D matrix.
```{r}
# generate 51 observations
observations <- obs_(51)
# plug the n*1 matrix into the dissimilarity martix function
D <- D_dis(observations)
# select a part of the dissimilarity matrix
D[1:5,1:5]
```  
This dissimilarity matrix looks good. Then I run the classical multidimensional scaling to this resulting matrix. Note, this is a non-Euclidean case. Some of the eigenvalue may be negative.
```{r}
d_mds <- cmdscale(D, k=50, eig = T)
lam <- d_mds$eig
# d_mds$points
cumsum(abs(lam))/sum(abs(lam))
cumsum(abs(lam^2))/sum(abs(lam^2))
```  
These values suggest that the first two coordinates will give an adequate representation of the simulated dissimilarity distances. Then, I make the scatter plot using the first two scores.
```{r, eval=FALSE, included=FALSE}
lim <- range(d_mds$points[,1] * (-1)) * 1.2

plot(d_mds$points[,1]*(-1), d_mds$points[,2]*(-1), 
     xlab = "Coordinate 1", ylab = "Coordinate 2",
     xlim = lim, ylim = lim)
```  
```{r}
lim <- range(d_mds$points[,1]) * 1.2
plot(d_mds$points[,1], d_mds$points[,2], 
     xlab = "Coordinate 1", ylab = "Coordinate 2",
     xlim = lim, ylim = lim)
text(d_mds$points[,1], d_mds$points[,2], labels = c(1:51), pos=2,cex = 0.7)
```  
This two-dimensional plot has the symmetrical shape, which is reasonable since the mid-point,i.e., the 26th point, has the same dissimilarity with both the points on its left and right. Therefore, if we project all point on y-axis, it actually represent the dissimilarity degree between the each point and the 26th point.

## Ex. 4.2  
*Write an R function to calculate the chi-squared distance matrices for both rows and columns in a two-dimensional contingency table.*  

**MY SOLUTION**
Based on the definition of chi-squared distance matrices, I write the function as follows:
```{r, eval=FALSE, include=FALSE}
# note! This code chunk wont run and display in the output pdf!
# this code chunk is written by chatGPT!!
chi_squared_dist_matrices <- function(tbl) {
  
  # Calculate row sums and column sums
  row_sums <- apply(tbl, 1, sum)
  col_sums <- apply(tbl, 2, sum)
  
  # Calculate total count and expected counts
  total_count <- sum(tbl)
  exp_counts <- outer(row_sums, col_sums) / total_count
  
  function(x) sum((x - exp_counts[,match(col_sums[which(x > 0)], col_sums)])^2 / exp_counts[,match(col_sums[which(x > 0)], col_sums)])
  
  # Calculate chi-squared distances for rows and columns
  row_chi_sqdists <- apply(tbl, 1, function(x) sum((x - exp_counts[match(row_sums[which(x > 0)], row_sums),])^2 / exp_counts[match(row_sums[which(x > 0)], row_sums),]))
  col_chi_sqdists <- apply(tbl, 2, function(x) sum((x - exp_counts[,match(col_sums[which(x > 0)], col_sums)])^2 / exp_counts[,match(col_sums[which(x > 0)], col_sums)]))
  
  # Create matrices from row and column chi-squared distances
  row_chi_sqmat <- as.matrix(dist(rbind(row_chi_sqdists)))
  col_chi_sqmat <- as.matrix(dist(rbind(col_chi_sqdists)))
  
  # Return list of row and column chi-squared distance matrices
  return(list(row_chi_sqmat, col_chi_sqmat))
}

```
```{r}
chi_squared_dist_matrices <- function(tbl){
  # Calculate row sums and column sums
  row_sums <- apply(tbl, 1, sum)
  col_sums <- apply(tbl, 2, sum)
  # Calculate total count and expected counts
  total_count <- sum(tbl)
  
  # to have the dim of the input matrix
  c <- ncol(tbl)
  r <- nrow(tbl)
  # make a empty matrix to load all elements
  col_d_matrix <- matrix(0,c,c)
  
  # using loop to write the each ij entries into this matrix
  for (i in c(1:c)) {
    for (j in c(1:c)) {
      d_ij <- 0
      for (k in c(1:r)) {
        p_k_dot <- row_sums[k]/total_count
        p_k_i <- tbl[k,i]/col_sums[i]
        p_k_j <- tbl[k,j]/col_sums[j]
        d_ij <- d_ij + (1/p_k_dot)*(p_k_i-p_k_j)^2
      }
      col_d_matrix[i,j] <- d_ij
    }
  }
  # make a empty matrix to load all elements
  row_d_matrix <- matrix(0,r,r)
  
  # using loop to write the each ij entries into this matrix
  for (i in c(1:r)) {
    for (j in c(1:r)) {
      d_ij <- 0
      for (k in c(1:r)) {
        p_dot_k <- col_sums[k]/total_count
        p_i_k <- tbl[i,k]/row_sums[i]
        p_j_k <- tbl[j,k]/row_sums[j]
        d_ij <- d_ij + (1/p_dot_k)*(p_i_k-p_j_k)^2
      }
      row_d_matrix[i,j] <- d_ij
    }
  }
  return(list(col_d_matrix, row_d_matrix))
}
```  
This function return a list that contains both columns distance matrix and rows distance matrix. By using the index, like `[[1]]`or `[[2]]`, to extract the columns or rows distance matrix, respectively. Next, I created a two-dimensional contigency table to test this function.
```{r}
# Create a contingency table
tbl <- table(c("A", "A", "B", "B"), c("X", "Y", "X", "Z"))
tbl
# plug this simulated table into the function above
dist_m <-chi_squared_dist_matrices(tbl)
# the columns distance matrix is
dist_m[[1]]
# the row distance matrix is
dist_m[[2]]
```
It looks good!  






